Supplementary Figures: Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness

### 2017-06-02 manuscript/supp.tex
final tweaks on figs

\usepackage[font={small}]{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{lineno}
%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

\bibliographystyle{ieeetr}

\usepackage{bibunits}

@@

%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
\author{Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@

\section*{Abstract}
{\small
We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term robustness of salmon metapopulations.
}
\section{Introduction}

Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population vary across a metapopulation. 
This asynchrony will increase the potential for demographic rescue \cite{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}. 
For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \cite{Abbott:2017hl}. 
This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}. 
Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \cite{Thibaut:2012km}.
In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction.

It is clear that dispersal (i.e., movement with consequences for gene flow) can influence the evolutionary dynamics of the metapopulation, the interplay between selection, population dynamics, and persistence is not well understood.
Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \cite{MilnerGulland:2011vm}. 
Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \cite{Bell:2011ki,Carlson:2014is}.
% First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
Third, straying will influence the evolutionary dynamics of the metapopulation.
% though how the interplay between selection and population dynamics might impact persistence is less well understood. 
Although the dispersal of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \cite{Muhlfeld:2014hs}. 
More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \cite{Lowe:2015ft}.
Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
% Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.

%Movement through space (link back to straying)
% Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site. %; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}.
Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
Recently, the role of social interactions and collective navigation has been hypothesized [This issue].
The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
% Whether and to what extent the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
% How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.

%Eco-evolutionary dynamics in space
%% From geographic mosaic of coevolution to...
% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.

%Motivation: straying
The eco-evolutionary impacts of dispersal likely has important implications for conservation and management in key taxa such as in migratory salmon.
While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
These strays can introduce new maladaptive genotypes into the recipient population.
Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \cite{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
Thus, there is opportunity and a need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.

% Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
% However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
% Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
% Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.  
%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.

% Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
% These strays can introduce new maladaptive genotypes into the recipient population.
% Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}. 
% Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}.

% 
% \begin{figure}
%   \captionsetup{justification=raggedright,
% singlelinecheck=false
% }
% \centering
% \includegraphics[width=0.4\textwidth]{figs2/fig_xkcd.png}
% \caption{
% \emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
% A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
% In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
% Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
% } \label{fig:xkcd}
% \end{figure}

%Alternative stable state
Here we seek to explore how collective density-dependent straying influences the stability and robustness of metapopulations through ecological and evolutionary processes.
% the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
% Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance. 
This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs. the positive effects of demographic and evolutionary rescue.
% [We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]

\section{Model Description \& Analysis}

\noindent{\bf (a) Metapopulation framework}\\
\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determing recruitment rates.
We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.

%PW: Can we say anything about the effect of the recipient population size and that effect on evolutionary dynamics? Again, straying can be thought of as both from the donor or recipient population perspective.
The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
\caption{
A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$. Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
FB marks the fold bifurcation.
} \label{fig:traj}
\end{figure}

We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
In this sense, both populations serve as donor and recipient populations.
We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
This mix of individuals is subject to the same compensatory effects, which is determined by the parameter $\beta$.
Taken together, the difference equation that determine changes in population size is

\begin{align}
  &N_i(t+1) = \\ \nonumber
  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ \left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right) \\ \nonumber
  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
  \label{eq:N}
\end{align}

\noindent where the difference equation for $N_j$ mirrors that for $N_i$.

The recruitment of local individuals $(1-m)N_i(t)$ as a function of their mean trait value at time $t$ and the local trait optimum, is

\begin{align}
  &R_i[\mu_i(t)] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P},
  \label{eq:R}
\end{align}

\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
Recent studies of wild sockeye have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
\\

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm.pdf}
\caption{
(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
The black line denotes the fold bifurcation.
} \label{fig:PE}
\end{figure*}

% \noindent{\bf (b) Recruitment over a selective landscape}
\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where 
\begin{equation}
w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
\end{equation}
We make two simplifying assumptions.
First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \cite{Jasper:2013cc}.
Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.

An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation

\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
\end{align}

\noindent where the first two factors determine the mixed normal average of the aggregated local and remote populations.
The partial derivative in the Eq. \ref{eq:mu} determines how the mean trait changes through time due to natural selection \cite{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
\\

\noindent {\bf (b) Density-dependent straying}
%Density dependent m
\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is

\begin{equation}
  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
  \label{eq:ddm}
\end{equation}

\noindent where $C$ is a half-saturation constant.
We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
Thus, for realistic population densities, $m(t) < m_0$.\\

\noindent {\bf (c) Habitat heterogeneity}

%Linking m/m0 and thetadiff
The straying rate is largely influenced by the spatial distance between the donor and recipient population.
The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).\\

\noindent {\bf (d) Measuring metapopulation robustness}
\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \cite{Anderson:2013gb}, such that

\begin{equation}
\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
\label{eq:pe}
\end{equation}

\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
Portfolio effects greater than unity corresponds to less synchronization  \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.

Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
For example, if populations settle to alternative stable states (alternative stable states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
\\

\section{Results}

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
\caption{
(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium. The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
FB marks the fold bifurcation.
} \label{fig:thetaPE}
\end{figure}

%Alternative stable states

\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig. \ref{fig:traj}).
This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum). 
Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
\caption{
Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
FB marks the fold bifurcation.
} \label{fig:relax}
\end{figure*}

Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
The fold bifurcation (the black line in Figs. \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.

%Critical slowing down
As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).

% \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
% Return time as a measure of system persistence and relation to PE
As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.

Straying had non-linear impacts on the recovery time of populations. 
When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.\\

\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
%Portfolio effects
If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq. \ref{eq:ddm}).
%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.

%Recovery times
Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
%Recovery times
%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.5\textwidth]{figs2/fig_traitdiff.pdf}
\caption{
Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$. The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime. 
} \label{fig:traitdiff}
\end{figure}

>>>>

\usepackage[font={small}]{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{xr}
\externaldocument[extfig-]{msv4}
%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
% \usepackage{lineno}

%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

\linespread{1.5}

\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;}

\bibliographystyle{prsb}

% \usepackage{bibunits}

@@

% \Blindtext
%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Supplementary Figures: Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
\author{
Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@
